Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-05 paper.tex
Minor tweaks to notation

\begin{figure}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG generated under the SMC, represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. By removing recombinant nodes, we limit the information on the time to each
of these events to the windows (blue shaded array) delimited by by the age of
the child node and the minimum age of the parents of the edges connecting to that child node.
Blue (unary) nodes represent recombinant nodes. The blue arrows show the
familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
}
\label{fig:smc-unary}
\end{figure}

An ARG can therefore be thought of either as a collection of inheritance
relationships between ancestral haplotypes,
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $A$
inherited the left half of its genome from ancestor $g$
and the right half from ancestor $f$
is seen in the fact that the first two trees have $A$ below $g$,
and the remaining have $A$ below $f$.
The blue dotted line from $A$ in the second tree to $f$ in the third tree
draws our attention to the fact that
the path traced by $A$'s ancestry switches at that point.
In fact, $A$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labeled by black numbers) are either samples
or \emph{common ancestry} nodes --
ancestors that are the most recent common ancestor for two samples
at some place along the genome.

@@

These locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014},
\tsinfer{} \citep{kelleher_inferring_2019},
\kwarg{} \citep{ignatieva_kwarg_2021}, and
\texttt{SINGER} \citep{deng2024robust} estimate ARGs 
including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

@@

% coalescent introduction: backwards-in-time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
% 
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm 
to simulate the coalescent with
recombination was formulated backwards in time
by Hudson \citep{hudson_properties_1983}. 
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process 
\citep{wiuf_recombination_1999}.
Such an along-the-genome approach is considerably more complex as the next 
tree to be simulated depends on all previous trees.
This idea however provided the basis of the 
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards-in-time, 
as well as along-the-genome.
In its backwards-in-time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping} 
genomic intervals may coalesce. 
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of
the ARG, the Markovian property can be formulated as follows: 
the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be a coalescent node or it already has been:
looking at the second tree (Fig.\ref{fig:smc-unary}),
we know node $f$ will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, this is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
As such, we only require a single half-open interval to describe
the ancestral
material associated with each common ancestor node in the ARG.

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
\textit{how} coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards in time.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each lineage $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Since we look backwards in time, $t=0$ is today, and so
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

@@

Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
definition of the strict total order it follows that
if $I_{jk} = 1$ then $I_{kj} = 0$.

For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify the likelihood computations.
In particular,
although recombination affects \emph{which} lineages can coalesce into $j$,
it does not change the \emph{number} of such lineages:
in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.

@@

To ensure that the ancestral material carried by a single lineage can at
all time be represented by a single half-open interval, we only update the
endpoints of the involved lineages following common ancestry or recombination events
such that only the minimal half-open interval whose endpoints have not fully coalesced
is used to update $L(t)$. Note that empty intervals are not added to the state.
%By reasoning about the process in this way we can keep representing 
% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
% and the process terminates.

@@

The structure of the ARG, as a collection of \emph{edges} $E$ and \emph{nodes} $N$,
as well as the definition of both recombination and common ancestry events, allows us
to cleanly factor the likelihood of the ARG across its edges and nodes.
Not only does recombination
naturally only affect one lineage at a time without even depending
on the state of other lineages. Common ancestry events can be equally
associated with a single edge given definition \ref{def:coal}.
The next paragraphs detail precisely how the number and the nature of the realized events
can be determined based on the number of
parents and children of each of the nodes in $N$ as well as how the total hazard can be
factorized into the individual contributions of each of the edges.

More specifically,
let $\deg_o(n) = \vert\{e \in E : p_e=n\}\vert$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ common ancestry events,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Note however that this description is general and holds for the case of
more-than-binary mergers. Indeed, inferred ARGs (e.g.\ \texttt{tsinfer}) may
contain polytomies. By treating these as if there were zero-length edges
between a sequence of mergers; the order of these unresolved mergers does 
not affect the likelihood. Under the SMC the likelihood of a zero-length edge is zero.
This is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.

Similarly,
let $\deg_i(n) = \vert\{e \in E : c_e=n\}\vert$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

@@

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$ (see fig.\ \ref{fig:likelihood}).
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligbile links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occur, then the number of links not
affected by recombination is 1 less (see eq.\ \ref{eq:depth}).

Finally, we turn to the hazard associated with common ancestry events.
To do this,
we need only consider what happens along the left side of each edge.
Recall that we define a total order on edges,
by ordering by left endpoint and breaking ties by child
(i.e., $(p,c,x,y) < (p',c',x',y')$
if $x<x'$, or if $x=x'$ and $c<c'$).
The instantaneous rate of coalescence of a lineage whose left endpoint is at $x$
at a given time is equal to $\lambda$ multiplied
by the number of earlier lineages that overlap $x$ at that time.
Using this total order, we can define
$I_e(t)$ to be the total number of earlier edges that overlap edge $e$
and are present at time $t$.
Then, the cumulative hazard for common ancestry events of edge $e$
between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(u) du$.
If the edge $e = (p,c,x,y)$
is the leftmost segment of material inherited by node $c$
(e.g., the left hand segment ancestral to node $c$
in fig.\ ~\ref{fig:likelihood}),
then no recombination occurred along this edge and 
this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
However, if the edge is \emph{not} the leftmost segment
(as shown in fig.\ \ref{fig:likelihood}),
then it was initiated by a new recombination breakpoint,
and we need to integrate over possible times the breakpoint occurred.
If this is the case, there is another edge $(p',c,x',x)$ with the same child node,
immediately adjacent to the focal edge $(p,c,x,y)$.
The recombination event that split the two must have happened between $t_c$
and the smaller of $t_p$ and $t_{p'}$. Note that two events must have happened
between $t_c$ and $\min(t_p, t_{p'})$; taking all this into consideration,
the contribution to the likelihood here is
\begin{equation}\label{eq:depth}
B_e(r, \lambda) = \begin{cases}
    e^{-\lambda F(e, t_c, t_p)}
        & \qquad \text{if } x=x_{c} \\
    \int_{t_c}^{t_{p} \wedge t_{p'}} e^{- r s} e^{-\lambda F(e, s, t_{p})} ds
        & \qquad \text{if } x>x_{c} ,
\end{cases}
\end{equation}
where $x_c$ is the leftmost position of the edges whose children are $c$.

>>>>

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG generated under the SMC, represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. By removing recombinant nodes, we limit the information on the time to each
of these events to the windows (blue shaded array) delimited by by the age of
the child node and the minimum age of the parents of the edges connecting to that child node.
Blue (unary) nodes represent recombinant nodes. The blue arrows show the
familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
}
\label{fig:smc-unary}
\end{figure}

An ARG can therefore be thought of either as a collection of inheritance
relationships between ancestral haplotypes,
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $A$
inherited the left half of its genome from ancestor $g$
and the right half from ancestor $f$
is seen in the fact that the first two trees have $A$ below $g$,
and the remaining have $A$ below $f$.
The blue dotted line from $A$ in the second tree to $f$ in the third tree
draws our attention to the fact that
the path traced by $A$'s ancestry switches at that point.
In fact, $A$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labeled by black numbers) are either samples
or \emph{common ancestry} nodes --
ancestors that are the most recent common ancestor for two 
or more samples
at some place along the genome.

@@

These locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014},
\tsinfer{} \citep{kelleher_inferring_2019},
\kwarg{} \citep{ignatieva_kwarg_2021}, and
\texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

@@

% coalescent introduction: backwards-in-time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
% 
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm 
to simulate the coalescent with
recombination was formulated backwards in time
by Hudson \citeyearpar{hudson_properties_1983}. 
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process 
\citep{wiuf_recombination_1999}.
Such an along-the-genome approach is considerably more complex as the next 
tree to be simulated depends on all previous trees.
This idea however provided the basis of the 
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards-in-time, 
as well as along-the-genome.
In its backwards-in-time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping} 
genomic intervals may coalesce. (The SMC' additionally allows 
lineages with abutting genomic intervals to coalesce.)
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
rasmussen_genome-wide_2014}.
Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of
the ARG, the Markovian property can be formulated as follows: 
the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be (in the left-to-right sense) a coalescent node 
or it already has been:
looking at the second tree in Fig.\ref{fig:smc-unary},
we know node $f$ will be coalescent further along the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, this is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
As such, we only require a single half-open interval to describe
the ancestral
material associated with each common ancestor node in the ARG.

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards in time.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each lineage $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Since we look backwards in time, $t=0$ is today, and so
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

@@

Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
From the definition of the strict total order it follows that if $I_{jk} = 1$
then $I_{kj} = 0$. For later use, also define $C_j(t)$ to be the set of
lineages that can coalesce into lineage $j$, i.e., 
$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$; 
thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify the likelihood computations. In particular, although
recombination affects \emph{which} lineages can coalesce into $j$, it does not
change the \emph{number} of such lineages: in other words, a recombination
event changes $C_j(t)$ but not $I_{j}(t)$.

@@

[Do we need to get into this above-the-root subtelty here?]
To ensure that the ancestral material carried by a single lineage can at
all time be represented by a single half-open interval, we only update the
endpoints of the involved lineages following common ancestry or recombination events
such that only the minimal half-open interval whose endpoints have not fully coalesced
is used to update $L(t)$. Note that empty intervals are not added to the state.
%By reasoning about the process in this way we can keep representing 
% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
% and the process terminates.

@@

The structure of the ARG, as a collection of \emph{edges} $E$ and \emph{nodes} $N$,
as well as the definition of both recombination and common ancestry events, allows us
to cleanly factor the likelihood of the ARG across its edges and nodes.
Not only does recombination
naturally only affect one lineage at a time without even depending
on the state of other lineages. Common ancestry events can be equally
associated with a single edge given definition \eqref{def:coal}.
The next paragraphs detail precisely how the number and the nature of the realized events
can be determined based on the number of
parents and children of each of the nodes in $N$ as well as how the total hazard can be
factorized into the individual contributions of each of the edges.

More specifically,
let $\deg_o(u) = \vert\{e \in E : p_e=u\}\vert$ be the out-degree of a node $u$,
i.e., the number of edges having $u$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(u) - 1$ common ancestry events,
and so this contributes a factor of $\lambda^{\deg_o(u)-1}$ to the likelihood
for nodes with any children.
Note however that this description is general and holds for the case of
more-than-binary mergers. Indeed, inferred ARGs (e.g.\ \texttt{tsinfer}) may
contain polytomies. By treating these as if there were zero-length edges
between a sequence of mergers, the order of these unresolved mergers does 
not affect the likelihood. Under the SMC the likelihood of a zero-length edge is zero.
This is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.

Similarly,
let $\deg_i(u) = \vert\{e \in E : c_e=u\}\vert$ be the in-degree of a node $u$,
i.e., the number of edges having $u$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(u)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{u \in N} \lambda^{\deg_o(u)-1} r^{\deg_i(u)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

@@

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$ (see Fig.~\ref{fig:likelihood}).
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligible links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occur, then the number of links not
affected by recombination is 1 less (see eq.~\eqref{eq:depth}).

Finally, we turn to the hazard associated with common ancestry events.
To do this,
we need only consider what happens along the left side of each edge.
Recall that we define a total order on edges,
by ordering by left endpoint and breaking ties by child
(i.e., $(p,c,x,y) < (p',c',x',y')$
if $x<x'$, or if $x=x'$ and $c<c'$).
The instantaneous rate of coalescence of a lineage whose left endpoint is at $x$
at a given time is equal to $\lambda$ multiplied
by the number of earlier lineages that overlap $x$ at that time.
Using this total order, we can define
$I_e(t)$ to be the total number of earlier edges that overlap edge $e$
and are present at time $t$.
Then, the cumulative hazard for common ancestry events of edge $e$
between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$.
If the edge $e = (p,c,x,y)$
is the leftmost segment of material inherited by node $c$
(e.g., the left hand segment ancestral to node $c$
in fig.\ ~\ref{fig:likelihood}),
then no recombination occurred along this edge and 
this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
However, if the edge is \emph{not} the leftmost segment
(as shown in Fig.~\ref{fig:likelihood}),
then it was initiated by a new recombination breakpoint,
and we need to integrate over possible times the breakpoint occurred.
If this is the case, there is another edge $(p',c,x',x)$ with the same child node,
immediately adjacent to the focal edge $(p,c,x,y)$.
The recombination event that split the two must have happened between $t_c$
and the smaller of $t_p$ and $t_{p'}$. Note that two events must have happened
between $t_c$ and $\min(t_p, t_{p'})$; taking all this into consideration,
the contribution to the likelihood here is
\begin{equation}\label{eq:depth}
B_e(r, \lambda) = \begin{cases}
    e^{-\lambda F(e, t_c, t_p)}
        & \qquad \text{if } x=x_{c} \\
    \int_{t_c}^{t_{p} \wedge t_{p'}} e^{- r s} e^{-\lambda F(e, s, t_{p})} ds
        & \qquad \text{if } x>x_{c} ,
\end{cases}
\end{equation}
where $x_c$ is the leftmost position of the edges whose children are $c$.
